---
title: "Assignment 1 - Language Development in ASD - part 2"
author: "Riccardo Fusaroli"
date: "July 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Language development in Autism Spectrum Disorder (ASD)

Background: Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail: i) relying on actual naturalistic language production, ii) over extended periods of time. We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

We then want to test the language trajectory of child and parent over time.

This Markdown is structured in the following way:

1. The exercises: read them carefully. Under each exercise you will have to write your answers, once you have written and run the code. This is the part that you have to directly send to the teachers.
2. An (optional) guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results under the exercise part.
3. In exercise 4 you will be asked to create the best possible model of language development in TD and ASD children, picking and choosing whatever additional variables you want from the dataset. Next time, the models produced by the different groups will compete against each other to see who can produce the best model, so choose carefully!

You will have to have a github repository for the code and send the answers to Celine and Riccardo without code (but a link to your github repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

N.B. The following lines are a summary of the questions to be answered, the step-by-step instructions and tips are below.

## Exercise 1) Preliminary Data Exploration

Describe the participant samples in the dataset (e.g. by diagnosis, age, etc.). Do you think the two groups are well balanced? If not, what do you think was the reason? 

In order to investigate the balance between the two groups we performed t-tests in the form of linear models. We found significant differences between the two groups on Autism Diagnostic Observation Schedule (ADOS), which makes good sense as it is a measure of the severety of the ASD symptoms. We also found a significant difference in the age of the two groups of children. The ASD children were on average older than the TD children, but the reason for this difference between the two groups is probably that unlike TD children, ASD children vary a lot more in their language development. Some ASD children develop language faster than TD children, and some later. Futher more we found a significant difference between the two groups in the mean lenght of utterances (mlu) of the parents of the children. This is interesting because we did not find any significant difference in the mlu of the children in either groups. 

### Exercise 2) Children learning language: the effects of time and ASD
Describe linguistic development in TD and ASD children in terms of Mean Length of Utterance (MLU)?

We used R (R Core Team (2017)), and lme4 (Bates, Maechler, Bolker &amp; Walker, 2015), MuMIn
(Barton, 2016), and lmerTest (Kuznetsova, Brockhoff &amp; Christensen, 2016) in order to perform a linear
mixed effects analysis of the relationship between the participants' word production and diagnosis of ASD. As fixed effects, we chose to look into visit and diagnosis (with interaction). 
For random effects, we had by-subject random slopes and intercepts for visits.
Results showed that visit is significant as a main effect, although diagnosis is not. However the interaction between the two is significant for the childrens' mlu (β = 0.25, se = 0.038, t-value = 6.715, p < 0.001)
The full model with the fixed and random effects explained 82% of the variance in child mlu, while the fixed effects accounted for 35% of this variance. 

(note for me: If we make a model that vary both in intercept and slope, then we are less likely to be tricked by any misrepresentative “mean” value)


### Exercise 3) Child directed speech as a moving target
Describe how parental use of language changes over time in terms of MLU. What do you think is going on?

We used R (R Core Team (2017)), and lme4 (Bates, Maechler, Bolker &amp; Walker, 2015), MuMIn
(Barton, 2016), and lmerTest (Kuznetsova, Brockhoff &amp; Christensen, 2016) in order to perform a linear
mixed effects analysis of the relationship between the parents' mlu and their child's diagnosis. As fixed effects, we chose to look into visit and diagnosis (with interaction). 
For random effects, we had by-subject random slopes and intercepts for visits.
Results showed that both visit and diagnosis are significant as main effects, although the interaction between the two isn't significant for the parents' mlu (β = 0.040, se = 0.037, t-value = 1.096, p = 0.278)
The full model with the fixed and random effects explained 68% of the variance in parent mlu, while the fixed effects accounted for 23% of this variance. 

### Exercise 4) Looking into "individual differences" (demographic, clinical or cognitive profiles)
The dataset contains some additional variables characterizing the kids’ cognitive and clinical profile: ADOS (autism severity), MSEL EL (Expressive Language, that is, verbal IQ, or linguistic skills at first visit as assessed by a psychologist using Mullen Scales of Early Learning), MSEL VR (Visual Reception, used as a proxy for non verbal IQ at first visit), Age, Gender, Ethnicity. Would it make sense to add any of them to your model of linguistic trajectories? Create the best possible model (the one that best explain the data, with MLU as outcome). Next time your model will be tested on new participants, and we will proclaim a winner. Describe your strategy to select the best models (how did you choose the variables to include?) and send the code to Riccardo and Celine.

[HERE GOES YOUR ANSWER]

### [OPTIONAL] Exercise 5) Comment on how the three linguistic variables measure linguistic performance (the so-called "construct validity" of the measures). Do they express the same variance?

[HERE GOES YOUR ANSWER]

### Structure of the code chunks

Basic stuff:
- Loading the libraries
- Setting the directory and loading the data
- Look at the data (which variables are there? Are they in the right format?) and describe the participants (by diagnosis)

We will try to answer three questions:

- Do children with ASD develop language differently from non-ASD children?
- Do parents speak differently to children with ASD than to non-ASD ones?
- Which variables should we use to best explain the child linguistic performance?
  
### Loading the relevant libraries

Load necessary libraries : what will you need?

- e.g. something to plot with
- e.g. mixed effects models

```{r Load Libraries}

# loading packages
library(pacman)
p_load(plyr, dplyr, stringr, tidyverse, tidyr, ggplot2, pastecs, lmerTest, MuMIn, lme4)

```

### Define your working directory and load the data

- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data}
# setting working directory and loading data
getwd()
locpath= getwd()
setwd(locpath)
data = read.csv("~/Desktop/Exp. Met. 3/Task_1/Task_1_data.csv")

str(data)

```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Visit, Number of words used, Number of unique words used, length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r}
# Exercise 1
# Characterizing the participants

# subset with observations only from 1st visit
data_1 = subset(data, data$VISIT == "1")

# number of participants in each category of diagnosis
length(data[data$Diagnosis == 'ASD',1])
length(data[data$Diagnosis == 'TD',1])

# AGE
# lm test (in this case the same as a t-test) 
age = lm(Age ~ Diagnosis, data = data_1)
summary(age)
# significant = groups are not well matched. TD children are on average 12 months younger than the ASD children. 

# barplot
ggplot(age_data,aes(Diagnosis, Age)) +
 geom_bar(stat="summary",fun.y=mean)

# ETHNICITY
# GLM test 
ethnicity = glm(Diagnosis ~ Ethnicity, family = binomial, data_1)
summary(ethnicity)
# insignificant = groups are well matched 

# ADOS
ados = lm(ADOS_1 ~ Diagnosis, data = data_1)
summary(ados)
# significant = groups are not well matched. TD children on average score 13 points lower on the ADOS scale (severity of ASD)

# Verbal_IQ
verbal_IQ = lm(Verbal_IQ_1 ~ Diagnosis, data = data_1)
summary(verbal_IQ)
# insignificant = groups are well matched.

# Non_verbal_IQ
non_verbal_IQ = lm(Non_verbal_IQ_1 ~ Diagnosis, data = data_1)
summary(non_verbal_IQ)
# insignificant = groups are well matched.

# No. of words used overall (tokens_CHI)
overall = lm(tokens_CHI ~ Diagnosis, data = data_1)
summary(overall)
# insignificant = groups are well matched.

# No. of unique words used (types_CHI)
unique = lm(types_CHI ~ Diagnosis, data = data_1)
summary(unique)
# insignificant = groups are well matched.

# MLU Child
CHI_MLU = lm(CHI_MLU ~ Diagnosis, data = data_1)
summary(CHI_MLU)
# insignificant = groups are well matched.

# MLU Mother/parents 
MOT_MLU = lm(MOT_MLU ~ Diagnosis, data = data_1)
summary(MOT_MLU)
# significant = groups are not well matched. TD children's parents have on average longer utterances than ASD children's parents.

```

[REPORT THE RESULTS]

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r}

# graph of Child MLU over time (age)
ggplot(data, aes(data$Age, data$CHI_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Age of child (months)", y = "Child MLU", title = "Child MLU over time (age)") # + geom_jitter() 
# geom_jitter  It adds a small amount of random variation to the location of each point

# graph of child MLU over time (visits)
ggplot(data, aes(data$VISIT, data$CHI_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Visit number", y = "Child MLU", title = "Child MLU over visits")

# null model (mlu child predicted by intercept + random effects)
# (baseline model)
null = lmer(CHI_MLU ~ 1 + (1 + VISIT | SUBJ), data)
summary(null)

# mixed effects linear model (predicted by visit)
visit = lmer(CHI_MLU ~ 1 + VISIT + (1 + VISIT | SUBJ), data)
summary(visit)

# visit + diagnosis
visit_diagnosis= lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1 + VISIT | SUBJ), data)
summary(visit_diagnosis)

# visit * diagnosis (interaction. We use interaction to se if there is an interaction/difference between the slopes/development of the two groups, then we incorporate this difference in the model (not sure?))
visit_diagnosis_1= lmer(CHI_MLU ~ 1 + VISIT * Diagnosis + (1 + VISIT | SUBJ), data)
summary(visit_diagnosis_1)

# Best model?
anova(null, visit_diagnosis)

# Best model? (with interaction)
anova(null, visit_diagnosis_1)

# ususally we are interested in the marginal effects (the m value)
# the marginal values are those associated with the fixed effects
# the conditional values are those associated with the fixed effects + random effects
# this number explains how much of the variance is explained by the model (in this case 0.35 = 35%)
r.squaredGLMM(visit_diagnosis_1)

```

How would you evaluate whether the model is a good model?

```{r}



```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better

```{r}



```

Exciting right?
Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r}

ggplot(data, aes(data$VISIT, data$MOT_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Visit number", y = "Mother MLU", title = "Mother MLU over visits")

ggplot(data, aes(data$Age, data$MOT_MLU, colour = Diagnosis)) + geom_point() + geom_smooth(method = "lm") + labs(x = "Age of child (months)", y = "Mother MLU", title = "Mother MLU over time (age)")

# MOT visit + diagnosis
mot_visit_diagnosis= lmer(MOT_MLU ~ 1 + VISIT + Diagnosis + (1 + VISIT | SUBJ), data)
summary(mot_visit_diagnosis)

# MOT visit * diagnosis
mot_visit_diagnosis_1= lmer(MOT_MLU ~ 1 + VISIT * Diagnosis + (1 + VISIT | SUBJ), data)
summary(mot_visit_diagnosis_1)

# How much variance does the model explain?
r.squaredGLMM(mot_visit_diagnosis_1)

```

[REPORT THE RESULTS]

### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Celine.


```{r}



```

REPORT OF THE RESULTS